---
title: "IsoEM: EM-Based Transcript Quantification for Long-Read RNA-seq"
author: "hmutpw"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{IsoEM: Complete Workflow Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

**IsoEM** provides equivalence-class (EC) compressed EM quantification for
[IsoQuant](https://github.com/ablab/isoquant) long-read RNA-seq output.
It supports three analysis modes:

| Mode | Use case | Key input |
|------|----------|-----------|
| `bulk_single` | One sample, one BAM | `counts_file` + `sample_id` |
| `bulk_multi` | Joint IsoQuant run, multiple samples | `counts_file` + `anno_file` (read→sample) |
| `sc` | Single-cell / spatial (10x, Visium HD) | `counts_file` + `anno_file` (read→barcode+UMI) |

### Modular pipeline

```
prepare_isoem()  →  IsoEMInput     (validate paths, no data loaded)
       ↓
build_ec()       →  IsoEMEC        (bulk: build ECs)
build_sc_ec()    →  IsoEMEC        (SC:   build ECs with UMI dedup)
       ↓
run_em()         →  IsoEMResult / IsoEMDataset / IsoEMSCResult
       ↓
write_isoem()    /  write_sc_isoem()   (write results to disk)
```

One-step wrappers `run_isoem()` and `run_sc_isoem()` chain all steps
automatically for convenience.

---

## Installation

```r
# From GitHub
devtools::install_github("hmutpw/IsoEM")

# From local tarball
install.packages("IsoEM_0.3.0.tar.gz", repos = NULL, type = "source")
```

---

## Toy data

All examples use built-in toy datasets:

```{r toy_data}
library(IsoEM)

# Core input files
counts_f <- system.file("extdata", "toy_counts.tsv",       package = "IsoEM")
gtf_f    <- system.file("extdata", "toy.gtf",              package = "IsoEM")

# Single-cell annotation: read_id | barcode | umi
bc_f     <- system.file("extdata", "toy_bc_umi.tsv",       package = "IsoEM")

# Multi-sample: joint counts + read-to-sample mapping
multi_f  <- system.file("extdata", "toy_counts_multi.tsv", package = "IsoEM")
sid_f    <- system.file("extdata", "toy_sample_ids.tsv",   package = "IsoEM")

# Regex mode: read_id has embedded bc/umi (_bc=BARCODE_sq=UMI)
regex_f  <- system.file("extdata", "toy_counts_regex.tsv", package = "IsoEM")
```

Preview the files:

```{r preview}
# Counts file: read_id | transcript_id
head(read.table(counts_f, sep = "\t", header = FALSE,
                col.names = c("read_id", "transcript_id")), 4)

# SC annotation: read_id | barcode | umi
head(read.table(bc_f, sep = "\t", header = FALSE,
                col.names = c("read_id", "barcode", "umi")), 4)

# Regex counts: read_id encodes barcode + UMI
head(read.table(regex_f, sep = "\t", header = FALSE,
                col.names = c("read_id", "transcript_id")), 3)
```

---

## Workflow 1: Bulk single-sample

### One-step

```{r bulk_single_onestep}
result <- run_isoem(
  counts_file = counts_f,
  gtf_file    = gtf_f,
  mode        = "bulk_single",
  sample_id   = "toy_sample",
  outdir      = file.path(tempdir(), "isoem_bulk"),
  verbose     = FALSE
)
print(result)
```

### Step-by-step (modular)

```{r bulk_single_modular}
# Step 1: validate inputs — returns lightweight IsoEMInput object
input <- prepare_isoem(
  counts_file = counts_f,
  gtf_file    = gtf_f,
  mode        = "bulk_single",
  sample_id   = "toy_sample",
  outdir      = file.path(tempdir(), "isoem_bulk2"),
  verbose     = FALSE
)
print(input)

# Step 2: build equivalence classes
ec <- build_ec(input, verbose = FALSE)
print(ec)

# Step 3: run EM
result <- run_em(ec, gtf_file = gtf_f, verbose = FALSE)
print(result)

# Inspect counts table
head(result$counts[, .(transcript_id, gene_id, em_count, tpm, certainty)])
```

### Write results

```{r bulk_single_write, eval=FALSE}
write_isoem(result, outdir = "results/toy_sample/")
# Output:
#   results/toy_sample/counts.tsv   — transcript-level EM counts
#   results/toy_sample/qc.tsv       — run QC summary
```

---

## Workflow 2: Bulk multi-sample (joint IsoQuant run)

```{r bulk_multi}
# Step 1: prepare — anno_file maps read_id → sample_id
input_multi <- prepare_isoem(
  counts_file = multi_f,
  gtf_file    = gtf_f,
  mode        = "bulk_multi",
  anno_file   = sid_f,       # read_id | sample_id
  outdir      = file.path(tempdir(), "isoem_multi"),
  verbose     = FALSE
)
print(input_multi)

# Step 2 + 3: EC + EM
ec_multi     <- build_ec(input_multi, verbose = FALSE)
result_multi <- run_em(ec_multi, gtf_file = gtf_f, verbose = FALSE)
print(result_multi)

# Access per-sample results
head(result_multi$count_matrix[1:5, ])
```

---

## Workflow 3: Single-cell (anno_file mode)

The standard SC workflow uses a pre-prepared `read_id | barcode | umi`
file (e.g. extracted from a minimap2 BAM using the bundled shell script).

### One-step

```{r sc_onestep}
result_sc <- run_sc_isoem(
  counts_file = counts_f,
  gtf_file    = gtf_f,
  anno_file   = bc_f,
  unit        = "umi",       # UMI-level quantification (recommended)
  outdir      = file.path(tempdir(), "isoem_sc"),
  verbose     = FALSE
)
print(result_sc)
```

### Step-by-step

```{r sc_modular}
# Step 1: validate
input_sc <- prepare_isoem(
  counts_file  = counts_f,
  gtf_file     = gtf_f,
  mode         = "sc",
  anno_file    = bc_f,
  unit         = "umi",
  outdir       = file.path(tempdir(), "isoem_sc2"),
  verbose      = FALSE
)
print(input_sc)

# Step 2: build SC ECs (with UMI dedup)
ec_sc <- build_sc_ec(input_sc, verbose = FALSE)
print(ec_sc)

# Step 3: run EM
result_sc <- run_em(ec_sc, gtf_file = gtf_f, verbose = FALSE)
print(result_sc)

# Inspect cell QC
head(result_sc$cell_qc)
```

### With barcode whitelist

```{r sc_whitelist}
# Filter to known valid cell barcodes
valid_barcodes <- paste0("CELL0", 1:5)   # keep only cells 1-5

input_wl <- prepare_isoem(
  counts_file  = counts_f,
  gtf_file     = gtf_f,
  mode         = "sc",
  anno_file    = bc_f,
  barcodes_use = valid_barcodes,
  unit         = "umi",
  outdir       = file.path(tempdir(), "isoem_sc_wl"),
  verbose      = FALSE
)

ec_wl     <- build_sc_ec(input_wl, verbose = FALSE)
result_wl <- run_em(ec_wl, gtf_file = gtf_f, verbose = FALSE)
cat("Cells after whitelist filter:", ncol(result_wl$count_matrix), "\n")
```

### Write SC results

```{r sc_write, eval=FALSE}
write_sc_isoem(result_sc, outdir = "results/sc_sample/")
# Output:
#   results/sc_sample/matrix/matrix.mtx    — sparse count matrix
#   results/sc_sample/matrix/features.tsv  — transcript annotations
#   results/sc_sample/matrix/barcodes.tsv  — cell barcodes
#   results/sc_sample/cell_qc.tsv          — per-cell QC
```

### Load into Seurat

```{r seurat, eval=FALSE}
library(Seurat)
# write_sc_isoem output is directly compatible with Read10X
seu <- Read10X("results/sc_sample/matrix/")
seu <- CreateSeuratObject(seu, project = "sc_sample")
```

---

## Workflow 4: Single-cell (regex extraction mode)

When barcode and UMI are embedded in the read ID (e.g. from BLAZE or
custom demultiplexing), use the `pattern` argument to extract them via
named capture groups without a separate annotation file.

```{r sc_regex}
# toy_counts_regex.tsv has read IDs like:
# READ_00001_bc=CELL03_sq=AGATCTAC
head(read.table(regex_f, sep = "\t", header = FALSE)[, 1], 3)

# Named capture groups: (?P<name>pattern)
input_regex <- prepare_isoem(
  counts_file = regex_f,
  gtf_file    = gtf_f,
  mode        = "sc",
  pattern     = ".*_bc=(?P<barcode>[A-Z0-9]+)_sq=(?P<umi>[ACGT]+)",
  unit        = "umi",
  outdir      = file.path(tempdir(), "isoem_regex"),
  verbose     = FALSE
)

ec_regex     <- build_sc_ec(input_regex, verbose = FALSE)
result_regex <- run_em(ec_regex, gtf_file = gtf_f, verbose = FALSE)
print(result_regex)
```

Alternatively, specify barcode and UMI patterns separately:

```{r sc_regex_sep, eval=FALSE}
input_sep <- prepare_isoem(
  counts_file = regex_f,
  gtf_file    = gtf_f,
  mode        = "sc",
  bc_pattern  = "_bc=([A-Z0-9]+)",    # first capture group = barcode
  umi_pattern = "_sq=([ACGT]+)",      # first capture group = UMI
  unit        = "umi",
  outdir      = file.path(tempdir(), "isoem_regex2"),
  verbose     = FALSE
)
```

---

## EC object: save and reuse

The `IsoEMEC` object can be saved and reused to re-run EM with different
parameters without re-processing the large input files:

```{r ec_reuse}
# Build EC once
ec <- build_sc_ec(
  prepare_isoem(counts_f, gtf_f, mode = "sc", anno_file = bc_f,
                unit = "umi", outdir = tempdir(), verbose = FALSE),
  verbose = FALSE
)

# Save EC object
ec_path <- file.path(tempdir(), "sc_ec.rds")
saveRDS(ec, ec_path)

# Later: load and re-run EM with different parameters
ec_loaded <- readRDS(ec_path)
result_tight <- run_em(ec_loaded, gtf_file = gtf_f,
                       max_iter = 1000, tol = 1e-8, verbose = FALSE)
print(result_tight)
```

---

## Extracting barcode/UMI from BAM

For users starting from a minimap2-aligned BAM, IsoEM provides a shell
script to extract `read_id | barcode | umi`:

```bash
# Default tags: BC:Z (barcode), U8:Z (UMI)
bash $(Rscript -e "cat(system.file('scripts',
  'extract_barcodes_from_bam.sh', package='IsoEM'))") \
    -i aligned.bam \
    -o read_bc_umi.tsv.gz \
    -t 8

# Custom tags (e.g. STARsolo / CellRanger output)
bash extract_barcodes_from_bam.sh \
    -i aligned.bam -o out.tsv.gz \
    -b CB:Z -u UB:Z
```

The output is a 3-column TSV used as `anno_file` in `prepare_isoem()`.

---

## Output format reference

### Bulk counts table (`counts.tsv`)

| Column | Description |
|--------|-------------|
| `transcript_id` | Transcript identifier |
| `sample_id` | Sample name |
| `em_count` | EM-estimated read count |
| `tpm` | Transcripts Per Million |
| `unique_count` | Reads uniquely assigned to this transcript |
| `total_count` | All reads compatible with this transcript |
| `certainty` | `unique_count / total_count` — quantification confidence (0–1) |
| `multimapping_rate` | `1 − certainty` |
| `gene_id` | Gene identifier (from GTF) |
| `is_novel` | Novel IsoQuant transcript |

### SC sparse matrix

Output follows the 10x Genomics / STARsolo Matrix Market convention:

```
outdir/
├── matrix/
│   ├── matrix.mtx[.gz]       # sparse (transcript × cell), EM counts
│   ├── features.tsv[.gz]     # transcript_id, gene_id, is_novel
│   └── barcodes.tsv[.gz]     # one barcode per line
└── cell_qc.tsv[.gz]          # per-cell QC metrics
```

### Cell QC table (`cell_qc.tsv`)

| Column | Description |
|--------|-------------|
| `barcode` | Cell barcode |
| `n_transcripts_detected` | Transcripts with EM count > 0.01 |
| `em_converged` | Whether EM converged |
| `em_n_iter` | Actual EM iterations |
| `n_ec` | Number of ECs for this cell |
| `n_unique_ec` | ECs with only one compatible transcript |

---

## Note on transposable elements

IsoEM was designed with TE-derived transcript quantification in mind.
Highly repetitive loci (e.g. L1 LINE elements) typically show
`certainty < 0.1` because a single read may be compatible with hundreds of
nearly identical copies. This is **expected behaviour**, not an error — the
EM algorithm distributes counts probabilistically across all compatible
copies. For downstream differential expression, filtering by `certainty`
threshold can help focus on high-confidence assignments.

---

## Session info

```{r session}
sessionInfo()
```
